ANTS: Open-Source Tools for Normalization And Neuroanatomy

### 2009-11-25 Documentation/Tex/antstheory.tex
final design for theory paper
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@215 fd2da0a4-d968-48df-9cb5-d8c496287cbf

%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
\documentclass[10pt,twocolumn]{IEEEtran}

\newcommand{\fv}{\mbox{\boldmath $f$}}
\newcommand{\Iv}{\mbox{\boldmath $I$}}
\newcommand{\xv}{\mbox{\boldmath $x$}}
\newcommand{\velocity}{\mbox{\boldmath $v$}}
\newcommand{\velocityinv}{\mbox{\boldmath $v$}^{-1}}
\newcommand{\Velocityinv}{\mbox{\boldmath $V$}^{-1}}
\newcommand{\Velocity}{\mbox{\boldmath $V$}}
\newcommand{\welocity}{\mbox{\boldmath $w$}}
\newcommand{\spatialimagegradient}{\mbox{\boldmath $I$}\mbox{$_{x}$}}
\newcommand{\perpspatialimagegradient}{\mbox{\boldmath $I$}\mbox{$^{\perp}_{x}$}}
\newcommand{\temporalimagederivative}{\mbox{$I_t$}}
\newcommand{\finitestrain}{\mbox{\boldmath $\varepsilon^{\ast}$}}
\newcommand{\smallstrain}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\dispgradient}{{\bf \nabla \displace}}
\newcommand{\displace}{{\bf u}} 
\newcommand{\Id}{\text{\bf Id}}
\newcommand{\ode}{{\em O.D.E.}}
\newcommand{\odes}{{\em O.D.E.}s}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\phiinv}{\phi^{-1}}
\newcommand{\psiinv}{\psi^{-1}}
\newcommand{\dM}{DM}
\newcommand{\DM}{Diffeomorphometry}
\newcommand{\diff}{diffeomorphism}
\newcommand{\Diff}{Diffeomorphism}
\newcommand{\orbit}{\mathcal{O}}
\newcommand{\avg}{\mathcal{A}}
\newcommand{\avgn}{\mathcal{A}_n}
\newcommand{\avgna}{\mathcal{A}^a_{n}}
\newcommand{\nset}{\{ J_i \}_n}
\newcommand{\bari}{\bar{I}}
\newcommand{\bart}{\bar{t}}
\newcommand{\jac}{\mathcal{J}}
\newcommand{\pert}{\mbox{\boldmath $w$}}
\newcommand{\barpert}{\bar{\mbox{\boldmath $w$}}}
\newcommand{\half}{0.5}

\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\Z}{{\bf Z}}
\newcommand{\z}{{\bf z}}
\newcommand{\p}{{\bf p}}
\newcommand{\Y}{{\bf Y}}
\newcommand{\y}{{\bf y}}
\newcommand{\ytild}{\tilde{{\bf y}}}
\newcommand{\q}{{\bf q}}
\newcommand{\surf}{\mathcal{S}}
\newcommand{\phij}{\phi_{ij}}
\newcommand{\aphij}{\bar{\phi}_{j}}
\newcommand{\apsij}{\bar{\psi}_{j}}
\newcommand{\aphi}{\bar{\phi}}
\newcommand{\aphiinv}{\bar{\phi}^{-1}}

\newcommand{\domain}{\Omega}
\newcommand{\meanshape}{\bf \bar{x}}
\newcommand{\g}{\mbox{\boldmath $g$}}
\newcommand{\h}{\mbox{\boldmath $h$}}
\newcommand{\yv}{\mbox{\boldmath $y$}}

@@

\newcommand{\evol}{{E_{\text{Vol}}}}
\newcommand{\epair}{{E_{\text{Pair}}}}
\newcommand{\esec}{{E_{S}}}
\newcommand{\inv}{^{-1}}
\newcommand{\vinit}{ \velocity^0_{ij}(\x,\tau=0) }
\newcommand{\avinit}{ \bar{\velocity}^0_{j}(\x,t) }
\newcommand{\avinitz}{ \bar{\velocity}^0_{j}(\x,t_a) }
\newcommand{\gz}{\nabla_{\z}}

\newcommand{\Section}[1]{\vspace{-8pt}\section{\hskip-1em.~~#1}\vspace{-3pt}}
\newcommand{\SubSection}[1]{\vspace{-3pt}\subsection{\hskip -1em.~~#1}\vspace{-3pt}}

\newcommand{\BASE}{.}%/Users/stnava/Work/}
\newcommand{\TEXDIR}{.}%{\BASE Words/Writing/}
\newcommand{\ROOT}{.}%{\BASE Words/}
\newcommand{\FIGDIR}{.}%{\BASE Words/Writing/MEDIAITK/wbir/}
\newcommand{\FIGDIRAIBS}{.}%{\BASE Words/Writing/IPMI2003/jpegs/}
\newcommand{\FIGDIRP}{proposal/}%{\BASE Words/Writing/proposal/}
\newcommand{\FIGDIRPJ}{proposal/jpegs/}%{\BASE Words/Writing/proposal/jpegs/}
\newcommand{\FIGDIRADAD}{.}%{\BASE Words/Writing/ADAD/}
\newcommand{\FIGS}{.}%{\BASE Words/Writing/neuroimageipam-final/FIGS/}

%% bare_jrnl.tex
%% V1.2
%% 2002/11/18
%% by Michael Shell
%% mshell@ece.gatech.edu
%%
%% NOTE: This text file uses MS Windows line feed conventions. When (human)
%% reading this file on other platforms, you may have to use a text
%% editor that can handle lines terminated by the MS Windows line feed
%% characters (0x0D 0x0A).
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.6b or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.ieee.org
%% and/or
%% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/
%%
%% This code is offered as-is - no warranty - user assumes all risk.
%% Free to use, distribute and modify.

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% Testflow can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/testflow

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print. Authors are encouraged to use U.S. letter paper when 
% submitting to IEEE. Use the testflow package mentioned above to verify
% correct handling of both paper sizes by the author's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
% This example can be formatted using the peerreview
% (instead of journal) mode.
%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
\usepackage{amssymb,amsmath}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{rotating}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{listing}						
\usepackage{listings}					
\graphicspath{./Figures/}

% If the IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it:
% \documentclass[journal]{../sty/IEEEtran}

% some very useful LaTeX packages include:

%\usepackage{cite}      % Written by Donald Arseneau
                        % V1.6 and later of IEEEtran pre-defines the format
                        % of the cite.sty package \cite{} output to follow
                        % that of IEEE. Loading the cite package will
                        % result in citation numbers being automatically
                        % sorted and properly "ranged". i.e.,
                        % [1], [9], [2], [7], [5], [6]
                        % (without using cite.sty)
                        % will become:
                        % [1], [2], [5]--[7], [9] (using cite.sty)
                        % cite.sty's \cite will automatically add leading
                        % space, if needed. Use cite.sty's noadjust option
                        % (cite.sty V3.8 and later) if you want to turn this
                        % off. cite.sty is already installed on most LaTeX
                        % systems. The latest version can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/cite/

%\usepackage{graphicx}  % Written by David Carlisle and Sebastian Rahtz
                        % Required if you want graphics, photos, etc.
                        % graphicx.sty is already installed on most LaTeX
                        % systems. The latest version and documentation can
                        % be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/required/graphics/
                        % Another good source of documentation is "Using
                        % Imported Graphics in LaTeX2e" by Keith Reckdahl
                        % which can be found as esplatex.ps and epslatex.pdf
                        % at: http://www.ctan.org/tex-archive/info/
% NOTE: for dual use with latex and pdflatex, instead load graphicx like:
%\ifx\pdfoutput\undefined
%\usepackage{graphicx}
%\else
%\usepackage[pdftex]{graphicx}
%\fi

% However, be warned that pdflatex will require graphics to be in PDF
% (not EPS) format and will preclude the use of PostScript based LaTeX
% packages such as psfrag.sty and pstricks.sty. IEEE conferences typically
% allow PDF graphics (and hence pdfLaTeX). However, IEEE journals do not
% (yet) allow image formats other than EPS or TIFF. Therefore, authors of
% journal papers should use traditional LaTeX with EPS graphics.
%
% The path(s) to the graphics files can also be declared: e.g.,
% \graphicspath{{../eps/}{../ps/}}
% if the graphics files are not located in the same directory as the
% .tex file. This can be done in each branch of the conditional above
% (after graphicx is loaded) to handle the EPS and PDF cases separately.
% In this way, full path information will not have to be specified in
% each \includegraphics command.
%
% Note that, when switching from latex to pdflatex and vice-versa, the new
% compiler will have to be run twice to clear some warnings.

%\usepackage{psfrag}    % Written by Craig Barratt, Michael C. Grant,
                        % and David Carlisle
                        % This package allows you to substitute LaTeX
                        % commands for text in imported EPS graphic files.
                        % In this way, LaTeX symbols can be placed into
                        % graphics that have been generated by other
                        % applications. You must use latex->dvips->ps2pdf
                        % workflow (not direct pdf output from pdflatex) if
                        % you wish to use this capability because it works
                        % via some PostScript tricks. Alternatively, the
                        % graphics could be processed as separate files via
                        % psfrag and dvips, then converted to PDF for
                        % inclusion in the main file which uses pdflatex.
                        % Docs are in "The PSfrag System" by Michael C. Grant
                        % and David Carlisle. There is also some information 
                        % about using psfrag in "Using Imported Graphics in
                        % LaTeX2e" by Keith Reckdahl which documents the
                        % graphicx package (see above). The psfrag package
                        % and documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/psfrag/

%\usepackage{subfigure} % Written by Steven Douglas Cochran
                        % This package makes it easy to put subfigures
                        % in your figures. i.e., "figure 1a and 1b"
                        % Docs are in "Using Imported Graphics in LaTeX2e"
                        % by Keith Reckdahl which also documents the graphicx
                        % package (see above). subfigure.sty is already
                        % installed on most LaTeX systems. The latest version
                        % and documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/subfigure/

%\usepackage{url}       % Written by Donald Arseneau
                        % Provides better support for handling and breaking
                        % URLs. url.sty is already installed on most LaTeX
                        % systems. The latest version can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/other/misc/
                        % Read the url.sty source comments for usage information.

%\usepackage{stfloats}  % Written by Sigitas Tolusis
                        % Gives LaTeX2e the ability to do double column
                        % floats at the bottom of the page as well as the top.
                        % (e.g., "\begin{figure*}[!b]" is not normally
                        % possible in LaTeX2e). This is an invasive package
                        % which rewrites many portions of the LaTeX2e output
                        % routines. It may not work with other packages that
                        % modify the LaTeX2e output routine and/or with other
                        % versions of LaTeX. The latest version and
                        % documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/sttools/
                        % Documentation is contained in the stfloats.sty
                        % comments as well as in the presfull.pdf file.
                        % Do not use the stfloats baselinefloat ability as
                        % IEEE does not allow \baselineskip to stretch.
                        % Authors submitting work to the IEEE should note
                        % that IEEE rarely uses double column equations and
                        % that authors should try to avoid such use.
                        % Do not be tempted to use the cuted.sty or
                        % midfloat.sty package (by the same author) as IEEE
                        % does not format its papers in such ways.

%\usepackage{amsmath}   % From the American Mathematical Society
                        % A popular package that provides many helpful commands
                        % for dealing with mathematics. Note that the AMSmath
                        % package sets \interdisplaylinepenalty to 10000 thus
                        % preventing page breaks from occurring within multiline
                        % equations. Use:
%\interdisplaylinepenalty=2500
                        % after loading amsmath to restore such page breaks
                        % as IEEEtran.cls normally does. amsmath.sty is already
                        % installed on most LaTeX systems. The latest version
                        % and documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% Other popular packages for formatting tables and equations include:

%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty which improves the
% LaTeX2e array and tabular environments to provide better appearances and
% additional user controls. array.sty is already installed on most systems.
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

% Mark Wooding's extremely powerful MDW tools, especially mdwmath.sty and
% mdwtab.sty which are used to format equations and tables, respectively.
% The MDWtools set is already installed on most LaTeX systems. The lastest
% version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/mdwtools/

% V1.6 of IEEEtran contains the IEEEeqnarray family of commands that can
% be used to generate multiline equations as well as matrices, tables, etc.

% Also of notable interest:

% Scott Pakin's eqparbox package for creating (automatically sized) equal
% width boxes. Available:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/eqparbox/

% Notes on hyperref:
% IEEEtran.cls attempts to be compliant with the hyperref package, written
% by Heiko Oberdiek and Sebastian Rahtz, which provides hyperlinks within
% a document as well as an index for PDF files (produced via pdflatex).
% However, it is a tad difficult to properly interface LaTeX classes and
% packages with this (necessarily) complex and invasive package. It is
% recommended that hyperref not be used for work that is to be submitted
% to the IEEE. Users who wish to use hyperref *must* ensure that their
% hyperref version is 6.72u or later *and* IEEEtran.cls is version 1.6b
% or later. The latest version of hyperref can be obtained at:
%
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/hyperref/
%
% Also, be aware that cite.sty (as of version 3.9, 11/2001) and hyperref.sty
% (as of version 6.72t, 2002/07/25) do not work optimally together.
% To mediate the differences between these two packages, IEEEtran.cls, as
% of v1.6b, predefines a command that fools hyperref into thinking that
% the natbib package is being used - causing it not to modify the existing
% citation commands, and allowing cite.sty to operate as normal. However,
% as a result, citation numbers will not be hyperlinked. Another side effect
% of this approach is that the natbib.sty package will not properly load
% under IEEEtran.cls. However, current versions of natbib are not capable
% of compressing and sorting citation numbers in IEEE's style - so this
% should not be an issue. If, for some strange reason, the user wants to
% load natbib.sty under IEEEtran.cls, the following code must be placed
% before natbib.sty can be loaded:
%
% \makeatletter
% \let\NAT@parse\undefined
% \makeatother
%
% Hyperref should be loaded differently depending on whether pdflatex
% or traditional latex is being used:
%
%\ifx\pdfoutput\undefined
%\usepackage[hypertex]{hyperref}
%\else
%\usepackage[pdftex,hypertexnames=false]{hyperref}
%\fi
%
% Pdflatex produces superior hyperref results and is the recommended
% compiler for such use.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

% The lineno packages adds line numbers. Start line numbering with
% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
% for the whole article with \linenumbers.
 \usepackage{lineno}

%\linenumbers
\begin{document}

%Setup listing environment
\definecolor{listcomment}{rgb}{0.0,0.5,0.0}
\definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
\definecolor{listbackground}{gray}{0.965}

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listbackground}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        tabsize = 2,
        language=[ISO]C++,
        floatplacement=!h
        }

%
% paper title
\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
\author{Brian B. Avants, Nicholas J. Tustison, Gang Song, and James C. Gee
%               Penn Image Computing and Science Laboratory \\
%               University of Pennsylvania \\
%               3600 Market Street, Suite 370 \\
%               Philadelphia,  PA  19104 \\
%               contact email:  tustison@grasp.cis.upenn.edu
%\thanks{Manuscript received XXX; revised XXX.
 %       }% <-this % stops a space
}

@@

\begin{abstract}
Computational anatomy (CA) seeks to quantify natural variation in
biological shape and function with roots that reach back to the
seminal works of Charles Darwin and D'arcy Thompson.  CA is currently
applied to study health, disease and epidemiology and uses deformable
mappings between images as a basic technique.  However, there is a
lack of standards and reproducibility in the field that is due, in
part, to the use of proprietary software and private data that is difficult 
to openly evaluate.  To
facilitate reproducibility in CA measurements and advancement of
imaging sciences, NIH has recently committed significant support to
open-source data and software resources.  Here, we report a recent
product of this commitment: Advanced Normalization Tools (ANTS), an
ITK-based toolkit for CA and related areas.  The ANTS open-source
library consists of a well-evaluated suite of state-of-the-art
normalization and template-building tools for quantitative
morphometric analysis.  We highlight the prominent features of ANTS
and demonstrate its utility by performing a detailed analysis on
openly-available anatomically labeled brain data from the non-rigid
image registration evaluation project (NIREP).  The results from this
analysis evidences the high level of accuracy achievable with ANTS
using intensity-based registration alone.  In addition, we show the
significant performance gains may be achieved by coupling
intensity-based image metrics and point-set-based metrics from
specific, sensibly selected cortical structures.
%Computational anatomy is the quantitative study of biological shape and its natural variations which traces its roots to the seminal works of Charles Darwin and D'arcy Thompson.  Integral to this research field is the calculation of deformable mappings between images that serve as invaluable measurement tools.  Technological evolution has spawned much innovation in the computational development of such deformable mapping algorithms.  However, not only is algorithmic development essential for increasing scientific knowledge but the availability of these technologies is equally crucial for facilitating the progress of the scientific community at large.  This is the motivation behind such projects as the Insight Toolkit (ITK) of the National Institutes of Health which serves as a repository for many popular image segmentation and registration routines.  In addition, the ITK library provides a well-structured framework for several stand-alone projects such as Elastix, ITK-SNAP, ManagedITK, the Medical Imaging Toolkit (MITK), WrapITK, and many others.  Our own  contribution to the field of computational anatomy, also built on an ITK foundation, is called Advanced Normalization Tools (ANTS).  The ANTS open-source library consists of an entire suite of state-of-the-art normalization and template-building tools for quantitative morphometric analysis. After highlighting the prominent features of ANTS, we demonstrate its utility by performing a detailed analysis on anatomically labeled brain data from the non-rigid image registration evaluation project (NIREP).  The results from this analysis evidences the high level of accuracy achievable with ANTS using intensity-based registration alone.  In addition, we show the performance gains achieved by coupling intensity-based image metrics and point-set-based metrics from specific, sensibly selected cortical structures.

%In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.
{\bf FIXME -- make clear what transformations we are comparing with constant similarity ... }
\end{abstract}

%\IEEEkeywords{
%ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
%}

%\IEEEpeerreviewmaketitle
\maketitle

\section{Introduction}

The rapid advancement of biological and medical imaging technologies
has caused a proliferation in the development of quantitative tools
for computational anatomy.  The principal tools of this emerging field
are deformable mappings between images whether they be driven by
similarity metrics which are intensity-based, point-set based, or
both.  Several categories of mappings have been proposed in the
literature.  Of particular recent interest are diffeomorphic
transformations which, by definition, preserve topology.  Topology
preservation is fundamental to making comparisons between objects in
the natural world that are thought to change in such a way that local
neighborhoods are preserved.  Cytoarchitectonic brain mapping studies
also suggest that the layout of cell types throughout the brain is
generally preserved \cite{Schleicher2009}, further motivating diffeomorphic mapping
in the context of the brain.

@@

We first provide an overview of the different components included in
ANTS, such as the available transformation models and similarity
metrics offered.  This is followed by an extensive experimental
analysis that builds upon the results from the recent image
normalization evaluation of \cite{Klein2009} which was limited to a
single configuration of ANTS.  This will provide a useful benchmark
for evaluating other possible ANTS configurations.  Importantly, this
work also suggests---to our knowledge, for the first time---the clear
practical benefit of time-parameterized diffeomorphic mapping over the
greedy and exponential mapping approaches.

\section{Theoretical Overview of ANTS}
A useful classification schema of normalization techniques is based upon the following three principal components \cite{Brown1992,Ibanez2002}:
\begin{itemize}
  \item the {\em transformation model},
  \item the {\em similarity (or correspondence) measures},  and
  \item the {\em optimization strategy}.
\end{itemize}
In general, image normalization is the process of finding the optimal transformation, $\phi$, within a specified transformation space which maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost function, $\mathcal{C}$, describing the similarity between $\mathcal{I}$ and $\mathcal{J}$, is minimized.   A summary of available transformation models and similarity measures are provided in Table \ref{table:chart}.  Details are given in subsequent sections.

\subsection{ANTS Transformation Models}

%\begin{figure*}
%  \centering
%    \begin{tabular}{c}
%      \includegraphics[width=170mm]{TransformationTable}
%    \end{tabular}
%  \caption{Available transformations and similarity metrics available in ANTS.  Similarity metric acronyms:  CC = cross correlation, MSQ = mean squares, MI = mutual information, JHCT = Jensen-Havrda-Charvat-Tsallis divergence, PSE = point-set expectation.  }
%  \label{fig:chart}
%\end{figure*}

>>>>

%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
\documentclass[10pt,twocolumn]{IEEEtran}
\newcommand{\fv}{\mbox{\boldmath $f$}}
\newcommand{\Iv}{\mbox{\boldmath $I$}}
\newcommand{\xv}{\mbox{\boldmath $x$}}
\newcommand{\velocity}{\mbox{\boldmath $v$}}
\newcommand{\velocityinv}{\mbox{\boldmath $v$}^{-1}}
\newcommand{\Velocityinv}{\mbox{\boldmath $V$}^{-1}}
\newcommand{\Velocity}{\mbox{\boldmath $V$}}
\newcommand{\welocity}{\mbox{\boldmath $w$}}
\newcommand{\spatialimagegradient}{\mbox{\boldmath $I$}\mbox{$_{x}$}}
\newcommand{\perpspatialimagegradient}{\mbox{\boldmath $I$}\mbox{$^{\perp}_{x}$}}
\newcommand{\temporalimagederivative}{\mbox{$I_t$}}
\newcommand{\finitestrain}{\mbox{\boldmath $\varepsilon^{\ast}$}}
\newcommand{\smallstrain}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\dispgradient}{{\bf \nabla \displace}}
\newcommand{\displace}{{\bf u}} 
\newcommand{\Id}{\text{\bf Id}}
\newcommand{\ode}{{\em O.D.E.}}
\newcommand{\odes}{{\em O.D.E.}s}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\phiinv}{\phi^{-1}}
\newcommand{\psiinv}{\psi^{-1}}
\newcommand{\dM}{DM}
\newcommand{\DM}{Diffeomorphometry}
\newcommand{\diff}{diffeomorphism}
\newcommand{\Diff}{Diffeomorphism}
\newcommand{\orbit}{\mathcal{O}}
\newcommand{\avg}{\mathcal{A}}
\newcommand{\avgn}{\mathcal{A}_n}
\newcommand{\avgna}{\mathcal{A}^a_{n}}
\newcommand{\nset}{\{ J_i \}_n}
\newcommand{\bari}{\bar{I}}
\newcommand{\bart}{\bar{t}}
\newcommand{\jac}{\mathcal{J}}
\newcommand{\pert}{\mbox{\boldmath $w$}}
\newcommand{\barpert}{\bar{\mbox{\boldmath $w$}}}
\newcommand{\half}{0.5}
\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\Z}{{\bf Z}}
\newcommand{\z}{{\bf z}}
\newcommand{\p}{{\bf p}}
\newcommand{\Y}{{\bf Y}}
\newcommand{\y}{{\bf y}}
\newcommand{\ytild}{\tilde{{\bf y}}}
\newcommand{\q}{{\bf q}}
\newcommand{\surf}{\mathcal{S}}
\newcommand{\phij}{\phi_{ij}}
\newcommand{\aphij}{\bar{\phi}_{j}}
\newcommand{\apsij}{\bar{\psi}_{j}}
\newcommand{\aphi}{\bar{\phi}}
\newcommand{\aphiinv}{\bar{\phi}^{-1}}
\newcommand{\domain}{\Omega}
\newcommand{\meanshape}{\bf \bar{x}}
\newcommand{\g}{\mbox{\boldmath $g$}}
\newcommand{\h}{\mbox{\boldmath $h$}}
\newcommand{\yv}{\mbox{\boldmath $y$}}

@@

\newcommand{\evol}{{E_{\text{Vol}}}}
\newcommand{\epair}{{E_{\text{Pair}}}}
\newcommand{\esec}{{E_{S}}}
\newcommand{\inv}{^{-1}}
\newcommand{\vinit}{ \velocity^0_{ij}(\x,\tau=0) }
\newcommand{\avinit}{ \bar{\velocity}^0_{j}(\x,t) }
\newcommand{\avinitz}{ \bar{\velocity}^0_{j}(\x,t_a) }
\newcommand{\gz}{\nabla_{\z}}
\newcommand{\Section}[1]{\vspace{-8pt}\section{\hskip-1em.~~#1}\vspace{-3pt}}
\newcommand{\SubSection}[1]{\vspace{-3pt}\subsection{\hskip -1em.~~#1}\vspace{-3pt}}
\newcommand{\BASE}{.}%/Users/stnava/Work/}
\newcommand{\TEXDIR}{.}%{\BASE Words/Writing/}
\newcommand{\ROOT}{.}%{\BASE Words/}
\newcommand{\FIGDIR}{.}%{\BASE Words/Writing/MEDIAITK/wbir/}
\newcommand{\FIGDIRAIBS}{.}%{\BASE Words/Writing/IPMI2003/jpegs/}
\newcommand{\FIGDIRP}{proposal/}%{\BASE Words/Writing/proposal/}
\newcommand{\FIGDIRPJ}{proposal/jpegs/}%{\BASE Words/Writing/proposal/jpegs/}
\newcommand{\FIGDIRADAD}{.}%{\BASE Words/Writing/ADAD/}
\newcommand{\FIGS}{.}%{\BASE Words/Writing/neuroimageipam-final/FIGS/}
%
% This example can be formatted using the peerreview
% (instead of journal) mode.
%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
\usepackage{amssymb,amsmath}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{rotating}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{listing}						
\usepackage{listings}					
\graphicspath{./Figures/}
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{lineno}
\begin{document}
\definecolor{listcomment}{rgb}{0.0,0.5,0.0}
\definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
\definecolor{listbackground}{gray}{0.965}
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listbackground}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        tabsize = 2,
        language=[ISO]C++,
        floatplacement=!h
        }

%
% paper title
\title{ANTS: Open-Source Tools for Normalization And Neuroanatomy}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
\author{Brian B. Avants, Nicholas J. Tustison, Gang Song, and James C. Gee
%               Penn Image Computing and Science Laboratory \\
%               University of Pennsylvania \\
%               3600 Market Street, Suite 370 \\
%               Philadelphia,  PA  19104 \\
%               contact email:  tustison@grasp.cis.upenn.edu
%\thanks{Manuscript received XXX; revised XXX.
 %       }% <-this % stops a space
}

@@

\begin{abstract}
Computational anatomy (CA) seeks to quantify natural variation in
biological shape and function with roots that reach back to the
seminal works of Charles Darwin and D'arcy Thompson.  CA is currently
applied to study health, disease and epidemiology and uses deformable
mappings between images as a basic technique.  However, there is a
lack of standards and reproducibility in the field that is due, in
part, to the use of proprietary software and private data that is
difficult to openly evaluate.  To facilitate reproducibility in CA
measurements and advancement of imaging sciences, NIH has recently
committed significant support to open-source data and software
resources.  Here, we report a recent product of this commitment:
Advanced Normalization Tools (ANTS), an ITK-based toolkit for CA and
related areas.  The ANTS open-source library consists of a
well-evaluated suite of state-of-the-art normalization, segmentation
and template-building tools for quantitative morphometric analysis.
We highlight the prominent features of ANTS, including diffeomorphic
normalization methods, and demonstrate its utility by performing a
detailed analysis on openly-available anatomically labeled brain data
from the non-rigid image registration evaluation project (NIREP).  The
results from this analysis evidences the high level of accuracy
achievable with ANTS using intensity-based registration and
segmentation.  In addition, we show the significant performance gains
may be achieved by coupling intensity-based image metrics and
label-based metrics from specific, sensibly selected cortical
structures.  Additional features are highlighted in the appendix. 
%In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.
%{\bf FIXME -- make clear what transformations we are comparing with constant similarity ... }
\end{abstract}
%\IEEEkeywords{
%ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
%}
%\IEEEpeerreviewmaketitle
\maketitle

\section{Introduction}
The rapid advancement of biological and medical imaging technologies
has caused a proliferation in the development of quantitative tools
for computational anatomy.  The principal tools of this emerging field
are deformable mappings between images whether they be driven by
similarity metrics which are intensity-based, point-set based, or
both.  Several categories of mappings have been proposed in the
literature.  Of particular recent interest are diffeomorphic
transformations which, by definition, preserve topology.  Topology
preservation is fundamental to making comparisons between objects in
the natural world that are thought to change in such a way that local
neighborhoods are preserved.  Cytoarchitectonic brain mapping studies
also suggest that the layout of cell types throughout the brain is
generally preserved \cite{Schleicher2009}, further motivating
diffeomorphic mapping in the context of the brain.

@@

We first provide an overview of the different components included in
ANTS, such as the available transformation models and similarity
metrics offered.  This is followed by an extensive experimental
analysis that builds upon the results from the recent image
normalization evaluation of \cite{Klein2009} which was limited to a
single configuration of ANTS.  A set of ANTS tools are evaluated on
the sixteen subjects from the public, expert-labeled neuroanatomy in
the NIREP dataset.  This will provide a useful benchmark for
evaluating other possible ANTS configurations.  Importantly, this work
also suggests---to our knowledge, for the first time---the clear
practical benefit of time-parameterized diffeomorphic mapping over the
greedy and exponential mapping approaches.

The paper is organized as follows.  Section~\ref{sec:theory} gives an
overview of the transformation models and similarity metrics available
in ANTS, along with optimization schemes.  Section~\ref{sec:imp}
connects the theory with the user interface.  Section~\ref{sec:expt},
first, illustrates the performance of different transformation models
on a pair of ``classic'' image registration examples.  Second, we
report results on a series of large-scale experiments that profile
ANTS registration schema against a cortical labeling problem using the
NIREP evaluation dataset.  Finally, we close with discussion of the
toolkit and our findings.

\section{Theoretical Overview of ANTS}
\label{sec:theory}
A useful classification schema of normalization techniques is based
upon the following three principal components
\cite{Brown1992,Ibanez2002}:
\begin{itemize}
  \item the {\em transformation model},
  \item the {\em similarity (or correspondence) measures},  and
  \item the {\em optimization strategy}.
\end{itemize}
In general, image normalization is the process of finding the optimal
transformation, $\phi$, within a specified transformation space which
maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a
location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost
function, $\mathcal{C}$, describing the similarity between
$\mathcal{I}$ and $\mathcal{J}$, is minimized.  A summary of available
transformation models and similarity measures are provided in Table
\ref{table:chart}.  Details are given in subsequent sections.
